Validation of GMEA

### 2023-09-15 figs/step_01c_splithcy.Rmd
split and analyse

agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    tselfcont <- t.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=tselfcont$p.value)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

@@

## Split data

@@

```{r,split1}

seed=100

set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
g1b <- setdiff(g1,g1s)

set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
g2b <- setdiff(g2,g2s)

set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
g3b <- setdiff(g3,g3s)

gxs <- c(g1s,g2s,g3s)
gxs <- gxs[order(gxs)]

gxb <- c(g1b,g2b,g3b)
gxb <- gxb[order(gxb)]

dess <- design[c(gxs),]
mxs <- mx[,c(gxs)]

desb <- design[c(gxb),]
mxb <- mx[,c(gxb)]

## Limma + GMEA

Using the S dataset, apply limma to conduct differential methylation analysis,
then GMEA.
GMEA has two steps:

@@

```{r,limmagmea1}

d <- model.matrix(~dess$age+ dess$sex + dess$groups)

fit.reduced <- lmFit(mxs,d)
fit.reduced <- eBayes(fit.reduced)
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma <- dma[order(dma$P.Value),]
dma <- dma[,c("UCSC_RefGene_Name","t")]

dmagg <- agg(dma,cores=8)
m <- dmagg$gme_res_df
m <- m[,"median",drop=FALSE]
tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
sig <- subset(tres,fdr<0.05)
ssig <- sig

Now repeat with the b set.

```{r,limmagmea2}

d <- model.matrix(~desb$age+ desb$sex + desb$groups)

fit.reduced <- lmFit(mxb,d)
fit.reduced <- eBayes(fit.reduced)
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma <- dma[order(dma$P.Value),]
dma <- dma[,c("UCSC_RefGene_Name","t")]

dmagg <- agg(dma,cores=8)
m <- dmagg$gme_res_df
m <- m[,"median",drop=FALSE]
tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
sig <- subset(tres,fdr<0.05)
bsig <- sig

@@

ssig_up <- rownames(subset(ssig,t_median>0))
ssig_dn <- rownames(subset(ssig,t_median<0))

bsig_up <- rownames(subset(bsig,t_median>0))
bsig_dn <- rownames(subset(bsig,t_median<0))

v1 <- list("s up"=ssig_up, "s dn"=ssig_dn,
  "b up"=bsig_up,"b dn"=bsig_dn)

plot(euler(v1),quantities = TRUE)

>>>>

agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    tselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=tselfcont$p.value)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- wilcox.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

@@

## Split, limma, GMEA

@@

Using the S and B datasets separately, apply limma to conduct differential methylation analysis,
then GMEA.
GMEA has two steps:

@@

```{r,splitanalysis}

split_analyse <- function(seed){
  set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
  g1b <- setdiff(g1,g1s)
  set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
  g2b <- setdiff(g2,g2s)
  set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
  g3b <- setdiff(g3,g3s)
  gxs <- c(g1s,g2s,g3s)
  gxs <- gxs[order(gxs)]
  gxb <- c(g1b,g2b,g3b)
  gxb <- gxb[order(gxb)]
  dess <- design[gxs,]
  mxs <- mx[,gxs]
  desb <- design[gxb,]
  mxb <- mx[,gxb]

  #s dataset
  d <- model.matrix(~dess$age+ dess$sex + dess$groups)
  fit.reduced <- lmFit(mxs,d)
  fit.reduced <- eBayes(fit.reduced)
  dm <- topTable(fit.reduced,coef=4, number = Inf)
  dma <- merge(myann,dm,by=0)
  dma <- dma[order(dma$P.Value),]
  dma <- dma[,c("UCSC_RefGene_Name","t")]
  dmagg <- agg(dma,cores=8)
  m <- dmagg$gme_res_df
  m <- m[,"median",drop=FALSE]
  tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
  sig <- subset(tres,fdr<0.01)
  ssig <- sig

  #b dataset
  d <- model.matrix(~desb$age+ desb$sex + desb$groups)
  fit.reduced <- lmFit(mxb,d)
  fit.reduced <- eBayes(fit.reduced)
  dm <- topTable(fit.reduced,coef=4, number = Inf)
  dma <- merge(myann,dm,by=0)
  dma <- dma[order(dma$P.Value),]
  dma <- dma[,c("UCSC_RefGene_Name","t")]
  dmagg <- agg(dma,cores=8)
  m <- dmagg$gme_res_df
  m <- m[,"median",drop=FALSE]
  tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
  sig <- subset(tres,fdr<0.01)
  bsig <- sig

  return(list("ssig"=ssig,"bsig"=bsig))
}

@@

ol <- function(res){
  ssig <- res$ssig
  bsig <- res$bsig
  ssig_up <- rownames(subset(ssig,t_median>0))
  ssig_dn <- rownames(subset(ssig,t_median<0))
  bsig_up <- rownames(subset(bsig,t_median>0))
  bsig_dn <- rownames(subset(bsig,t_median<0))
  v1 <- list("s up"=ssig_up, "s dn"=ssig_dn,
    "b up"=bsig_up,"b dn"=bsig_dn)
  plot(euler(v1),quantities = TRUE)
}

res <- split_analyse(seed=200)
ol(res=res)
